Bayesian Inference in Numerical Cognition: A Tutorial Using JASP

### 2019-12-09 jncTutorial/manuscript.tex
Linear regression theory

Like any other discipline within the psychological and behavioral sciences, researchers in numerical cognition rely on statistical inference, and hypothesis testing in particular, to verify theoretical claims against observed data. In recent years, Bayesian inference has become increasingly popular as an alternative to classical null hypothesis significance testing (NHST). Bayesian inference operates under a fundamentally different framework from NHST; instead of hard \enquote{accept/reject} decisions about a null hypothesis, Bayesian inference works by quantifying knowledge about competing hypotheses in light of observed data. Hypotheses that do a better job of predicting data will receive increased support, whereas hypotheses that don't predict the data well will receive decreased support.

@@

In this tutorial paper, we will introduce the reader to the basics of Bayesian inference through the lens of some classic, well-cited studies in numerical cognition. In three detailed examples illustrating \(t\)-tests, linear regression, and ANOVA, we will walk the reader through performing and interpreting the results of Bayesian analyses using the software package JASP (JASP Team, 2019). JASP is a free, open-source statistical software package offering the user the option of both Bayesian and frequentist analyses, all housed within a familiar, easy-to-use graphical user interface. Below, we will give a brief introduction to Bayesian hypothesis testing, followed by the three tutorial examples.

@@

Hypothesis testing can be viewed as a process for validating general claims based on a small sample of the general population. To illustrate, suppose that we are policy makers confronted with a new study programme which claims to increase the mathematical abilities of sixth graders in the United States. To ease the exposition we use a (fictitious) scale called the Scale for Advancing Mathematical Ability (SAMA) to measure mathematical ability. As policy makers we know that the population mean SAMA score is \(\mu_{0}=50\) and that the standard deviation is \(\sigma=10\) for the population of sixth graders that are taught using the standard materials in the US.

The superiority claim about the new study programme can be formalised as the working hypothesis \(\mathcal{H}_{1} : \mu > 50\). Here \(\mu\) refers to the \emph{population} mean of SAMA score of the same sixth graders, but taught using the new study programme instead. Such a treatment population does not (yet) exist and \(\mu\) cannot be measured, and is, therefore, unknown.

Nonetheless, we can infer the validity of the working hypothesis \(\mathcal{H}_{1}\) by running an experiment in which \(n=90\) sixth graders are taught with the new study programme. Suppose that these particular participants scored an average SAMA score of \(\bar{x} = 52.07\). Clearly, this sample mean is larger than the status quo population mean of \(50\), but can we assume that this finding generalises to the whole population? If the answer is \enquote{yes} and we declare the working hypothesis to be verified, then how confident would we be about such a decision? Below we recall how \(p\)-values are typically used to make hard \enquote{yes} and \enquote{no} decisions, and we remind the reader why a \(p\)-value does not quantify the (un)certainty of such decisions.

@@

To deduce how hypotheses, i.e., claims about the population such as \(\mu\), affect the data, such as \(\bar{x}\), we have to link the former to the latter with a distribution. For instance, let us suppose a normal distribution provides this link. The population parameters, the distribution that links them to the data, and the data together make up a (statistical) model, which allows for quantitative predictions. For instance, the claim that the treatment population is identitical to the current population of sixth graders implies the null model \(\mathcal{M}_{0}\) comprised of (1) the null hypothesis \(\mathcal{H}_{0}: \mu = 50\), (2) the assumption \(\sigma = 10\), and (3) normally distributed SAMA scores. When all three assumptions hold true, we can mathematically deduce that the \emph{potential} average SAMA score consisting of a group of \(n = 90\) participants is expected to fall within the interval \(47.93 = 50 - 1.96 \frac{10}{\sqrt{n}}\) and \(52.07 = 50 + 1.96 \frac{10}{\sqrt{n}}\) with 95\% chance.
\footnote{We use ``chance'' to refer to probabilistic statements about the data.}

@@

Under the null model \(\mathcal{M}_{0}\), the more extreme data are potential outcomes of \(\bar{x}\) larger than the observed sample mean of \(52.07\), but also lower than the unobserved possible sample mean of \(47.93 = 50 - (52.07-50)\). Under the null, and the normality assumptions in particular, we can deduce that the chance of seeing a potential \(\bar{x}\) in \((- \infty, 47.93]\) or \([52.07, \infty)\) is 5\%, thus, \(p=0.05\). Note that \(1 - p = 0.95\), which is the 95\% chance of the non-extreme event of having \(\bar{x}\) fall within the interval \([47.93, 52.07]\), if the null model holds true. We use this example to emphasise that \(p\)-values help us reason from the null model to (potential) data, but they do not allow us to make quantitative statements from the actually observed data about the hypotheses.

@@

Bayesian testing addresses these problems as follows. First, it compares the two hypotheses on equal footing by also operationalising the alternative model. For the case at hand, \(\mathcal{M}_{1}\) would be comprised of the assumptions (1') \(\mathcal{H}_{1}: \mu > 50\), (2) \(\sigma =10\) as before, and (3) the data being normally distributed as before. By doing so, violations of normality and the assumed standard deviation will penalize both models equally. Secondly, to reason from a particular observed outcome to the general population, it employs Bayes rule. To do so, we are required to express our uncertainty about the two hypotheses in terms of so-called \emph{prior model probabilities} before data observation.

Prior model probabilities help us contextualise the testing problem. Working as policy makers, we might have seen many claims of improved study programmes that did not work out well. For instance, past claims of improved study programmes led to null results four times out of five (i.e., we place 4-to-1 odds against the claim of improvement), we might set \(P( \mathcal{M}_{1}) = 0.2\) and \(P(\mathcal{M}_{0}) = 0.8\). On the other hand, the designers of the new study programme to increase SAMA scores might feel strongly about the benefits of the problem, and so might choose \(P( \mathcal{M}_{1}) = 0.7\) and \(P(\mathcal{M}_{0}) = 0.3\) instead. If \(\mathcal{M}_{1}\) and \(\mathcal{M}_{0}\) are the only models under consideration we have \(0 < P( \mathcal{M}_{0}) < 1\) and \(P( \mathcal{M}_{1})= 1 - P( \mathcal{M}_{0})\). The Bayes factor \(\textnormal{BF}_{10}(d)\) is a function of the observed data \(d\) that updates these prior model probabilities to posterior model probabilities, as follows:

@@

Note that the data only enter the formula via the Bayes factor. For instance, if \(\textnormal{BF}_{10}(d) = 6\), then the sceptic's probability of the superiority of the new study programme increases to 60\%, i.e., \(P(\mathcal{M}_{1} \mid d)=0.60\), and retains 40\% probability of for the null after data observation. On the other hand, the proponent of the new study programme would report a posterior model probability of \(P(\mathcal{M}_{1} \mid d)=0.93\) and \(P(\mathcal{M}_{0} \mid d)=0.07\). Note that the posterior model probabilities combine what we believed prior to data observation with the evidence of the data in terms of the Bayes factors. Both sceptic and the proponent could, if they wish, declare the new study programme to be superior to the status quo, but the sceptic would be much less confident than the proponent. The resulting posterior model probabilities can, thus, be used to complement a decision with a measure of uncertainty.

@@

Unlike the \(p\)-value, there are no universal standards for when a Bayes factor provides a sufficient amount of evidence. However, people have made recommendations in the past regarding sizes of Bayes factors, and these provide us with a helpful, but very rough guideline on its interpretation, when we do not know the context of the problem. In particular, we mention the recommendations of Jeffreys (1961), who proposed the following scheme\footnote{We advise readers not to get too hung up on the specific labels and breakpoints in this table, see also @giollaInpresswhat.}:

@@

In principle, the Bayes factor does not come with a specified decision rule. We believe that context made explicit using prior model probabilities by field experts are much more important, than overly generalised \enquote{rules} such \(p < 0.05\) or \(\textnormal{BF} _{10} > k\), where \(k\) is some arbitrarily chosen number.

@@

In this section, we describe a number of practical examples of using Bayesian inference in the context of numerical cognition. In the examples, each based on a well-cited paper in numerical cognition, we demonstrate a number of JASP's features using synthetic data sets, each generated in R using specific model assumptions that closely match the patterns of observed data described in the original papers. The data sets, as well as the R script used to generate the data sets, are available for download from \url{https://git.io/JeXui}.

@@

The Bayesian \(t\)-test is based on the priors \(\pi_{1}\), which puts an \enquote{uninformative} prior on the nuisance parameter \(\sigma\) (e.g., Ly, Marsman, Verhagen, Grasman, \& Wagenmakers, 2017) in conjunction with a Cauchy prior on the population effect size \(\delta = \mu / \sigma\), and the prior \(\pi_{0}\) with the same \enquote{uninformative} prior on the nuisance parameter \(\sigma\). With these priors, the resulting Bayes factor has a number of desirable properties: (1) \textbf{Measurement invariant.} The Bayes factor as a measure of evidence does not change if the units of measurements are changed (e.g., Rouder, Speckman, Sun, Morey, \& Iverson, 2009). For instance, if the data were to consist of temperature measurements, the Bayes factor would be the same regardless of the data being entered as Celsius or Fahrenheit. (2) \textbf{Predictively matched.} The Bayes factor reflects perfectly indifference, i.e., \(\textnormal{BF}_{10}=1\) when the data are completely uninformative. A Bayes factor of one implies that the posterior model probabilities equal the prior model probabilities, and nothing has been learned; see Jeffreys (1961), for further details. (3) \textbf{Information consistent.} For a sufficiently large enough sample with size \(n\) fixed, the Bayes factor provides overwhelmingly support for the alternative, whenever \(t\)-value goes to infinity; see Gronau, Ly, and Wagenmakers (n.d.), Ly, Verhagen, and Wagenmakers (2016) for further details. (4) \textbf{(Asymptotically) consistent.} As the Bayes factor takes in an increasing number of samples of data points generated under the null model, then as \(n\) grows, the Bayes factor \(\textnormal{BF}_{10}\) tends to zero to convey support for the data generating model \(\mathcal{M}_{0}\). On the other hand, if \(n\) grows with data generated under the alternative, the Bayes factor \(\textnormal{BF}_{10}\) grows unbounded; see for instance Liang, Paulo, Molina, Clyde, and Berger (2008). (5) \textbf{Robust to changes in the sample size.} With the uninformative prior on \(\sigma\), this Bayes factor allows for optional stopping, that is, early stopping when there is enough evidence for the alternative (e.g., Hendriksen, Heide, \& Grünwald, 2018) without inflating the type 1 error. On top of this, the Bayesian \(t\)-test is also allows for optional continuation, which implies that one can build on top of previous results; see for instance Faulkenberry (2019), Grünwald, Heide, and Koolen (2019), Ly, Etz, Marsman, and Wagenmakers (n.d.). Note that properties (4), (5), and (6) imply that the Bayesian \(t\)-test, first derived by Jeffreys (1948), also has good frequentist properties.

@@

We are now ready to perform a Bayesian analysis of the simulated Verguts and De Moor (2005) data. First, since we will separately analyze the \enquote{Choose smaller} and \enquote{Choose larger} groups, we will need to apply a \emph{filter} to the data set; specifically, we will first consider only the \enquote{Choose smaller} group. There are several ways to do this in JASP, but the easiest is the \enquote{click filter}\footnote{For a demonstration of the different methods of filtering in JASP, see https://jasp-stats.org/2018/06/27/how-to-filter-your-data-in-jasp/.} First, click the header of the \texttt{condition} column. To remove all of the \texttt{larger} trials from the analysis and retain only the \texttt{smaller} trials, we simply \enquote{turn off} the \texttt{larger} value by clicking on the check mark next to it. The checkmark will be replaced by a cross, and you will be able to see that the \texttt{larger} trials have been greyed out. The filter dialog can be closed by clicking on the X on the right side of the dialog box. Notice that a funnel icon will appear next to the \texttt{condition} column header to indicate that a filter is currently being applied. Any filters can be removed by re-opening the filter dialog and clicking the eraser icon.

To perform a Bayesian one-sample \(t\)-test, click on the \enquote{T-Tests} button in the ribbon at the top, and select \enquote{Bayesian One Sample T-Test}. There are two \(t\)-tests that we wish to perform, but for clarity, we will start with describing one in detail. Before beginning, however, we define our two competing hypotheses explicitly. Since the numerical distance effect is expected to be negative, we define \(\mathcal{H}_{0}:\delta = 0\) and \(\mathcal{H}_{-}:\delta < 0\), where the minus in the subscript reminds of that the alternative is one-sided negative.

@@

The Bayes factor is \(BF_{-0}=9.027\), which indicates that the observed data are approximately 9 times more likely under the alternative hypothesis \(\mathcal{H}_{1}\) than under the null hypothesis \(\mathcal{H}_{0}\). According to the convention of Jeffreys (1961), this is considered \enquote{moderate} evidence for \(\mathcal{H}_{1}\), which states that the true population effect \(\delta\) (from which our participants' regression slopes are drawn) is negative. A useful visual display of this Bayes factor is provided in the plot in the bottom of the Results pane. This probability wheel (or \enquote{pizza plot}) depicts the Bayes factor as an odds ratio of 9:027-to-1 in favor of \(\mathcal{H}_{1}\) over the null.\footnote{To simplify matters, we identify the hypotheses to the models and use the notation \( \mathcal{H}_{k} \) interchangeably with \( \mathcal{M}_{k} \) for the \( k \)th model.}

The bottom plot shows this in more detail. The dashed line depicts the aforementioned Cauchy prior (with a default scale/prior width of \(r=1/\sqrt{2} = 0.707\)) on the population effect size \(\delta\), but restricted to negative values in accordance to the alternative \(\mathcal{H}_{-}\). The solid line depicts the posterior density for \(\delta\). Under the assumption that the effect cannot be positive, this density provides a measure of uncertainty about the size of the population effect \(\delta\) \emph{after} observing the data. We can see in Figure \ref{fig:ttestBayes} that the more probable values of \(\delta\) are now centered around \(\delta = -0.5\). For the testing problem, one specific value worth exploring is \(\delta=0\). Notice that relative to the prior, the \(y\)-coordinate at \(\delta=0\) in the posterior is smaller than the \(y\)-coordinate at \(\delta=0\). This reflects the fact that our posterior belief that regarding \(\mathcal{H}_{0} : \delta = 0\) has decreased after data observation. In fact, it has decreased by a factor of approximately 9, which is the Bayes factor we obtained. This relationship is no accident; For the \(t\)-test the Bayes factor can be represented as the ratio of the prior and posterior on the population effect \(\delta\) evaluated at the null hypothesis \(\mathcal{H}_{0}: \delta = 0\). This fact is known as the Savage-Dickey density ratio (Wagenmakers, Lodewyckx, Kuriyal, \& Grasman, 2010).

@@

One last question to consider in our analysis is how sensitive the Bayes factor is to changes in the specification of the Cauchy prior on the population effect \(\delta\). Clicking the box for \enquote{Bayes factor robustness check} will help us to answer this question. Figure \ref{fig:ttestRobust} displays the result of this robustness check. As mentioned before, the default prior for \(\delta\) is a Cauchy distribution centered around \(0\) with scale (or \enquote{width}) \(r=0.707\). Figure \ref{fig:ttestRobust} depicts how the Bayes factor changes with varying values of \(r\). The plot represents the Bayes factor as a continuous function of \(r\), and the legend at the top lists the Bayes factor at four values: the value of \(r\) that produces the maximum Bayes factor, the original user-specified prior (\(r=0.707\)), a \enquote{wide} prior (\(r=1\)), and an \enquote{ultrawide} prior (\(r=\sqrt{2}=1.414\)). This plot should be interpreted holistically - across a wide range of values for the prior scale \(r\), the resulting Bayes factor ranges from 6.3 to 9, which implies that the result is quite robust.

@@

It is again important to note that the 95\% credible interval that is displayed in the figure is conditional on \(\mathcal{H}_{-}\). In other words, it only makes sense if \(\mathcal{H}_{-}\) is the true model. However, in this case, we have moderate evidence that \(\mathcal{H}_{0}\) is the true model, so we can ignore this credible interval. Conditioned on the null \(\mathcal{H}_{0}\) our best guess for size of the effect would simply be \(\delta = 0\). It is also worth mentioning that there might be some confusion present when taking this credible interval at face value. Some might argue that since the credible interval does not contain 0 (its right endpoint is -0.004), this is actually evidence for \(\mathcal{H}_{-}\). However, this is simply not true. The apparent paradox lies in how the credible interval is computed; in JASP the 95\% credible interval is computed as the \emph{central} 95\% of the mass of the posterior for \(\delta\). Since there is no posterior mass to the right of \(\delta=0\), this central portion of the posterior must necessarily be contained strictly to the left of \(\delta=0\), thus producing the right end point that is slightly less than \(\delta=0\). However, if we were to instead specify the credible interval as 95\% of the posterior distribution with \emph{highest density}, the resulting credible interval would contain \(\delta=0\). Currently, this functionality is not available in JASP, so great care must be taken in interpreting the credible interval for \(\delta\) based on restricted priors.

@@

@@

\begin{itemize}
\tightlist
\item
  General ideas about linear regression.
\item
  Testing for an effect being zero or not. Explain in terms of the example.
\item
  Count the number of multiple models
\item
  Write as a testing of means
\item
  Relate to \(t\)-tests
\item
  Some history (optional): The Bayesian \(t\)-test was extended to Bayes factors for linear regression by Zellner and Siow (1980), perfected by Liang et al. (2008) and Bayarri et al. (2012), popularised by Rouder and Morey (2012) in psychology, and extended further by Li and Clyde (2018). This extension is in hindsight straightforward, once the relationship between \(t\)-tests and regression is made precise.
\end{itemize}

Explain how Bayesian linear regression works. Probably focus on using uniform prior model probabilities; beta-binomial may be too difficult for inclusion in this paper.

>>>>

Like any other discipline within the psychological and behavioral sciences, researchers in numerical cognition rely on statistical inference, and hypothesis testing in particular, to verify theoretical claims against observed data. In recent years, Bayesian inference has become increasingly popular as an alternative to classical null hypothesis significance testing (NHST). Bayesian inference operates under a fundamentally different framework from NHST; instead of hard \enquote{accept/reject} decisions about a null hypothesis, Bayesian inference works by quantifying evidence about competing hypotheses in light of observed data. Hypotheses that do a better job of predicting data will receive increased support, whereas hypotheses that do not predict the data well will receive decreased support.

@@

In this tutorial paper, we will introduce the reader to the basics of Bayesian inference through the lens of some classic, well-cited studies in numerical cognition. In three detailed examples illustrating \(t\)-tests, linear regression, and ANOVA, we will walk the reader through performing and interpreting the results of Bayesian analyses using the software package JASP (JASP Team, 2019). JASP is a free, open-source statistical software package offering the user the option of both Bayesian and frequentist analyses, all housed within a familiar, easy-to-use graphical user interface. Below, we give a brief introduction to Bayesian hypothesis testing, followed by the three tutorial examples.

@@

Hypothesis testing can be viewed as a process of validating general claims based on a small sample of the general population. To illustrate, suppose that we are policy makers confronted with a new study program which claims to increase the mathematical abilities of sixth graders in the United States. To ease the exposition we use a (fictitious) scale called the Scale for Advancing Mathematical Ability (SAMA) to measure mathematical ability. As policy makers we know that the population mean SAMA score is \(\mu_{0}=50\) and that the standard deviation is \(\sigma=10\) for the population of sixth graders that are taught using the standard materials in the US.

The superiority claim about the new study programme can be formalized as the working hypothesis \(\mathcal{H}_{1} : \mu > 50\). Here \(\mu\) refers to the \emph{population} mean of SAMA score of the same sixth graders, but taught using the new study program instead. Such a treatment population does not (yet) exist and \(\mu\) cannot be measured, and is, therefore, unknown.

Nonetheless, we can infer the validity of the working hypothesis \(\mathcal{H}_{1}\) by running an experiment in which, say, \(n=90\) sixth graders are taught with the new study programme. Suppose that these particular participants scored an average SAMA score of \(\bar{y} = 52.07\). Clearly, this sample mean is larger than the status quo population mean of \(50\), but can we assume that this finding generalizes to the whole population? If the answer is \enquote{yes} and we declare the working hypothesis to be verified, then how confident would we be about such a decision? Below we recall how \(p\)-values are typically used to make hard \enquote{yes} and \enquote{no} decisions, and we remind the reader why a \(p\)-value does not quantify the (un)certainty of such decisions.

@@

To deduce how hypotheses, i.e., claims about the population such as \(\mu\), affect the data, such as \(\bar{y}\), we have to link the former to the latter with a distribution. For instance, let us suppose a normal distribution provides this link. The population parameters, the distribution that links them to the data, and the data together make up a (statistical) model, which allows for quantitative predictions. For instance, the claim that the treatment population is identitical to the current population of sixth graders implies the null model \(\mathcal{M}_{0}\) comprised of (1) the null hypothesis \(\mathcal{H}_{0}: \mu = 50\), (2) the assumption \(\sigma = 10\), and (3) normally distributed SAMA scores. When all three assumptions hold true, we can mathematically deduce that the \emph{potential} average SAMA score for a group of \(n = 90\) participants is expected to fall within the interval \(47.93 = 50 - 1.96 \frac{10}{\sqrt{n}}\) and \(52.07 = 50 + 1.96 \frac{10}{\sqrt{n}}\) with 95\% chance.
\footnote{We use ``chance'' to refer to probabilistic statements about the data.}

@@

Under the null model \(\mathcal{M}_{0}\), the more extreme data are potential outcomes of \(\bar{y}\) larger than the observed sample mean of \(52.07\), but also lower than the unobserved possible sample mean of \(47.93 = 50 - (52.07-50)\). Under the null, and the normality assumptions in particular, we can deduce that the chance of seeing a potential \(\bar{y}\) in \((- \infty, 47.93]\) or \([52.07, \infty)\) is 5\%, thus, \(p=0.05\). Note that \(1 - p = 0.95\), which is the 95\% chance of the non-extreme event of having \(\bar{y}\) fall within the interval \([47.93, 52.07]\), if the null model holds true. We use this example to emphasize that \(p\)-values help us reason from the null model to (potential) data, but they do not allow us to make quantitative statements from the actually observed data about the hypotheses.

@@

Bayesian testing addresses these problems as follows. First, it compares the two hypotheses on equal footing by also operationalizing the alternative model. For the case at hand, \(\mathcal{M}_{1}\) would be comprised of the assumptions (1') \(\mathcal{H}_{1}: \mu > 50\), (2) \(\sigma =10\) as before, and (3) the data being normally distributed as before. By doing so, violations of normality and the assumed standard deviation will penalize both models equally. Secondly, to reason from a particular observed outcome to the general population, it employs Bayes rule. To do so, we are required to express our uncertainty about the two hypotheses in terms of so-called \emph{prior model probabilities} before data observation.

Prior model probabilities help us contextualize the testing problem. Working as policy makers, we might have seen many claims of improved study programs that did not work out well. For instance, if past claims of improved study programmes led to null results four times out of five (i.e., we place 4-to-1 odds against the claim of improvement), then we might set \(P( \mathcal{M}_{1}) = 0.2\) and \(P(\mathcal{M}_{0}) = 0.8\). On the other hand, the designers of the new study program to increase SAMA scores might feel strongly about the benefits of their program, and so might choose \(P( \mathcal{M}_{1}) = 0.7\) and \(P(\mathcal{M}_{0}) = 0.3\) instead. If \(\mathcal{M}_{1}\) and \(\mathcal{M}_{0}\) are the only models under consideration we have \(0 < P( \mathcal{M}_{0}) < 1\) and \(P( \mathcal{M}_{1})= 1 - P( \mathcal{M}_{0})\). The Bayes factor \(\textnormal{BF}_{10}(d)\) is a function of the observed data \(d\) that updates these prior model probabilities to posterior model probabilities, as follows:

@@

Note that the data only enter the formula via the Bayes factor. For instance, if \(\textnormal{BF}_{10}(d) = 6\), then the sceptic's probability of the superiority of the new study program increases to 60\%, i.e., \(P(\mathcal{M}_{1} \mid d)=0.60\), and retains 40\% probability of for the null after data observation. On the other hand, the proponent of the new study program would report a posterior model probability of \(P(\mathcal{M}_{1} \mid d)=0.93\) and \(P(\mathcal{M}_{0} \mid d)=0.07\). Note that the posterior model probabilities combine what we believed prior to data observation with the evidence of the data in terms of the Bayes factors. Both sceptic and the proponent could, if they wish, declare the new study programme to be superior to the status quo, but the sceptic would be much less confident than the proponent. The resulting posterior model probabilities can, thus, be used to complement a decision with a measure of uncertainty.

@@

Unlike the \(p\)-value, there are no universal standards for when a Bayes factor provides a sufficient amount of evidence. However, people have made recommendations in the past regarding sizes of Bayes factors, and these provide us with a helpful, but very rough guideline on its interpretation, when we do not know the context of the problem. In particular, we mention the recommendations of Jeffreys (1961), who proposed the following scheme
\footnote{We advise readers not to get too hung up on the specific labels and breakpoints in this table, see also @giollaInpresswhat.}:

@@

In principle, the Bayes factor does not come with a specified decision rule. We believe that context made explicit using prior model probabilities by field experts are much more important, than overly generalized \enquote{rules} such \(p < 0.05\) or \(\textnormal{BF} _{10} > k\), where \(k\) is some arbitrarily chosen number.

@@

In this section, we describe a number of practical examples of using Bayesian inference in the context of numerical cognition. In the examples, each based on a well-cited paper in numerical cognition, we demonstrate a number of JASP's features using synthetic data sets, each generated in R using specific model assumptions that closely match the patterns of observed data described in the original papers.The data sets, as well as the R script used to generate the data sets, are available for download from \url{https://git.io/JeXui}.

@@

The Bayesian \(t\)-test is based on the priors \(\pi_{1}\), which puts an \enquote{uninformative} prior on the nuisance parameter \(\sigma\) (e.g., Ly, Marsman, Verhagen, Grasman, \& Wagenmakers, 2017) in conjunction with a Cauchy prior on the population effect size \(\delta = \mu / \sigma\), and the prior \(\pi_{0}\) with the same \enquote{uninformative} prior on the nuisance parameter \(\sigma\). With these priors, the resulting Bayes factor has a number of desirable properties: (1) \textbf{Measurement invariant.} The Bayes factor as a measure of evidence does not change if the units of measurements are changed (e.g., Rouder, Speckman, Sun, Morey, \& Iverson, 2009). For instance, if the data were to consist of temperature measurements, the Bayes factor would be the same regardless of the data being entered as Celsius or Fahrenheit. (2) \textbf{Predictively matched.} The Bayes factor reflects perfectly indifference, i.e., \(\textnormal{BF}_{10}=1\) when the data are completely uninformative. A Bayes factor of one implies that the posterior model probabilities equal the prior model probabilities, and nothing has been learned; see Jeffreys (1961), for further details. (3) \textbf{Information consistent.} For a sufficiently large enough sample size \(n\) fixed, the Bayes factor provides overwhelmingly support for the alternative, whenever the observed \(t\)-value goes to infinity; see Gronau, Ly, and Wagenmakers (n.d.), Ly, Verhagen, and Wagenmakers (2016) for further details. (4) \textbf{(Asymptotically) consistent.} As the Bayes factor takes in an increasing number of samples of data points generated under the null model, then as \(n\) grows, the Bayes factor \(\textnormal{BF}_{10}\) tends to zero to convey support for the data generating model \(\mathcal{M}_{0}\). On the other hand, if \(n\) grows with data generated under the alternative, the Bayes factor \(\textnormal{BF}_{10}\) grows unbounded; see for instance Liang, Paulo, Molina, Clyde, and Berger (2008). (5) \textbf{Robust to changes in the sample size.} With the uninformative prior on \(\sigma\), this Bayes factor allows for optional stopping, that is, early stopping when there is enough evidence for the alternative (e.g., Hendriksen, Heide, \& Grünwald, 2018) without inflating the type 1 error. On top of this, the Bayesian \(t\)-test also allows for optional continuation, which implies that one can build on top of previous results; see for instance Faulkenberry (2019), Grünwald, Heide, and Koolen (2019), Ly, Etz, Marsman, and Wagenmakers (n.d.). Note that properties (4), (5), and (6) imply that the Bayesian \(t\)-test, first derived by Jeffreys (1948), also has good frequentist properties.

@@

We are now ready to perform a Bayesian analysis of the simulated Verguts and De Moor (2005) data. First, since we will separately analyze the \enquote{Choose smaller} and \enquote{Choose larger} groups, we will need to apply a \emph{filter} to the data set; specifically, we will first consider only the \enquote{Choose smaller} group. There are several ways to do this in JASP, but the easiest is the \enquote{click filter}
\footnote{For a demonstration of the different methods of filtering in JASP, see https://jasp-stats.org/2018/06/27/how-to-filter-your-data-in-jasp/.}
First, click the header of the \texttt{condition} column. To remove all of the \texttt{larger} trials from the analysis and retain only the \texttt{smaller} trials, we simply \enquote{turn off} the \texttt{larger} value by clicking on the check mark next to it. The check mark will be replaced by a cross, and you will be able to see that the \texttt{larger} trials have been greyed out. The filter dialog can be closed by clicking on the X on the right side of the dialog box. Notice that a funnel icon will appear next to the \texttt{condition} column header to indicate that a filter is currently being applied. Any filters can be removed by re-opening the filter dialog and clicking the eraser icon.

To perform a Bayesian one-sample \(t\)-test, click on the \enquote{T-Tests} button in the ribbon at the top, and select \enquote{Bayesian One Sample T-Test}. There are two \(t\)-tests that we wish to perform, but for clarity, we will start with describing one in detail. Before beginning, however, we define our two competing hypotheses explicitly. Since the numerical distance effect is expected to be negative, we define \(\mathcal{H}_{0}:\delta = 0\) and \(\mathcal{H}_{-}:\delta < 0\), where the minus in the subscript reminds us that the alternative is one-sided negative.

@@

The Bayes factor is \(BF_{-0}=9.027\), which indicates that the observed data are approximately 9 times more likely under the alternative hypothesis \(\mathcal{H}_{1}\) than under the null hypothesis \(\mathcal{H}_{0}\). According to the convention of Jeffreys (1961), this is considered \enquote{moderate} evidence for \(\mathcal{H}_{1}\), which states that the true population effect \(\delta\) (from which our participants' regression slopes are drawn) is negative. A useful visual display of this Bayes factor is provided in the plot in the bottom of the Results pane. This probability wheel (or \enquote{pizza plot}) depicts the Bayes factor as an odds ratio of 9:027-to-1 in favor of \(\mathcal{H}_{1}\) over the null.
\footnote{To simplify matters, we identify the hypotheses to the models and use the notation \( \mathcal{H}_{k} \) interchangeably with \( \mathcal{M}_{k} \) for the \( k \)th model.}
The bottom plot shows this in more detail. The dashed line depicts the aforementioned Cauchy prior (with a default scale/prior width of \(r=1/\sqrt{2} = 0.707\)) on the population effect size \(\delta\), but restricted to negative values in accordance to the alternative \(\mathcal{H}_{-}\). The solid line depicts the posterior density for \(\delta\). Under the assumption that the effect cannot be positive, this density provides a measure of uncertainty about the size of the population effect \(\delta\) \emph{after} observing the data. We can see in Figure \ref{fig:ttestBayes} that the more probable values of \(\delta\) are now centered around \(\delta = -0.5\). For the testing problem, one specific value worth exploring is \(\delta=0\). Notice that relative to the prior, the \(y\)-coordinate at \(\delta=0\) in the posterior is smaller than the \(y\)-coordinate at \(\delta=0\). This reflects the fact that our posterior belief regarding \(\mathcal{H}_{0} : \delta = 0\) has decreased after data observation. In fact, it has decreased by a factor of approximately 9, which is the Bayes factor we obtained. This relationship is no accident; For the \(t\)-test the Bayes factor can be represented as the ratio of the prior and posterior on the population effect \(\delta\) evaluated at the null hypothesis \(\mathcal{H}_{0}: \delta = 0\). This fact is known as the Savage-Dickey density ratio (Wagenmakers, Lodewyckx, Kuriyal, \& Grasman, 2010).

@@

One last question to consider in our analysis is how sensitive the Bayes factor is to changes in the specification of the Cauchy prior on the population effect \(\delta\). Clicking the box for \enquote{Bayes factor robustness check} will help us to answer this question. Figure \ref{fig:ttestRobust} displays the result of this robustness check. As mentioned before, the default prior for \(\delta\) is a Cauchy distribution centered around \(0\) with scale (or \enquote{width}) \(r=0.707\). Figure \ref{fig:ttestRobust} depicts how the Bayes factor changes with varying values of \(r\). The plot represents the Bayes factor as a continuous function of \(r\), and the legend at the top lists the Bayes factor at four values: the value of \(r\) that produces the maximum Bayes factor, the original user-specified prior (\(r=0.707\)), a \enquote{wide} prior (\(r=1\)), and an \enquote{ultrawide} prior (\(r=\sqrt{2}=1.414\)). This plot should be interpreted holistically --across a wide range of values for the prior scale \(r\), the resulting Bayes factor ranges from 6.3 to 9, which implies that the result is quite robust.

@@

It is again important to note that the 95\% credible interval that is displayed in the figure is conditional on \(\mathcal{H}_{-}\). In other words, it only makes sense if \(\mathcal{H}_{-}\) is the true model. However, in this case, we have moderate evidence that \(\mathcal{H}_{0}\) is the true model, so we can ignore this credible interval. Conditioned on the null \(\mathcal{H}_{0}\) our best guess for size of the effect would simply be \(\delta = 0\). It is also worth mentioning that there might be some confusion present when to take this credible interval at face value. Some might argue that since the credible interval does not contain 0 (its right endpoint is -0.004), this is actually evidence for \(\mathcal{H}_{-}\). However, this is simply not true. The apparent paradox lies in how the credible interval is computed; in JASP the 95\% credible interval is computed as the \emph{central} 95\% of the mass of the posterior for \(\delta\). Since there is no posterior mass to the right of \(\delta=0\), this central portion of the posterior must necessarily be contained strictly to the left of \(\delta=0\), thus producing the right end point that is slightly less than \(\delta=0\). However, if we were to instead specify the credible interval as 95\% of the posterior distribution with \emph{highest density}, the resulting credible interval would contain \(\delta=0\). Currently, this functionality is not available in JASP, so great care must be taken in interpreting the credible interval for \(\delta\) based on restricted priors.

@@

The questions we would like to address here are (1) \enquote{Which out of these two covariates predict fluency score?}, and (2) \enquote{How large is the effect of each covariate \(y_{i}\)?}. Note that the first question can be answered by (a) neither symbolic nor nonsymbolic, (b) only symbolic, (c) only nonsymbolic, and (d) both symbolic and nonsymbolic distance effects predict fluency score. Equivalently, these answers correspond to the hypotheses that the regression coefficients (a') of both symbolic and nonsymbolic, (b') of nonsymbol is zero, (c') of symbolic, and (d') of neither is zero. Hence, the first question is the testing problem, whereas the second question deals with the estimation problem.

@@

\subsection{Linear regression models}

The technical details described here are included for completeness, but are not strictly necessary for the remainder of this section. In this subsection we elaborate on linear regression models, before we discuss the Bayesian implementation.

We write \(Y_{i}\) for the continous outcome variable of the \(i\)th participant, and \(x_{i1}, x_{i2}, \ldots, x_{ip}\) for the \(i\)th participant's covariates. For the example at hand, \(Y_{i}\) represents a participant's mathematical \enquote{fluency} score, and we have \(p=2\) covariates consisting of the \(i\)th participant's measurements symbolicDE, and nonsymbolicDE.
